Checking Big Suffix and LCP Arrays by Probabilistic Methods

### 2017-02-06 draft/draft.tex
2017/02/06

While the study for efficient construction of suffix and LCP arrays is evolving, the programs implementing the proposed algorithms are commonly provided ``as is", with the purpose only for the performance evaluation experiemnts of the articles where they are reported. That is, these programms give no guarantee that they have correctly implemented the proposed algorithms. The programs for recently proprosed algorithms are becoming much more complicated than before, casuing more difficulties for program verifying and debugging\footnote{In our studies before, we have exprienced problems caused by bugs of the exisitng programs.}. As a common practice, a suffix or LCP array checker is also provided for verifying the correctness of a constructed array. For example, such a checker is provided in the software packages eSAIS \cite{xxx} (DC3, pScan and etc? more and better) for constructing suffix and/or LCP arrays. In addition to help avoid implementation bugs, a checker is also demanded for an array constructed by a probabilistic algorithm \cite{xxx}. In this case, the array is correct with a probability and hence must be verified by a checker to ensure its correctness.

@@

@@

@@

		\If{$ST_1.{\sf notEmpty}()$ {\sf and} $ST_1.{\sf top}().1st = i$}{~\label{alg:1:h}

			$e := ST_1.{\sf top}()$, $ST_1.{\sf pop}()$, $e.3rd := fp$, $ST_1'.{\sf push}(e)$
		}
		\Else{

			\Return false \hspace{5cm} // condition (1) is violated
		}~\label{alg:1:i}

		\While{$ST_2.{\sf notEmpty}()$ {\sf and} $ST_2.{\sf top}().1st = i$}{

@@

	\Return true

@@

As will be seen from our experimental study in Section~\ref{sec:experiment}, Algorithm~\ref{alg:1} is quite space consuming, the peak disk use is 40 bytes per input character. This motivates us to design a lightweight checker using less space. In this section, we propose a method for checking the suffix and LCP arrays based on the induced-sorting principle. In comparison with Algorithm~\ref{alg:1}, the algorithm designed by this method only consumes about half space on real-world data sets.

\subsection{Preliminaries} \label{sec:algorithm2:preliminaries}

The induced sorting principle was previously employed to design algorithms on building suffix and LCP arrays in internal and external memory. The basic idea is to induce lexical order and LCP-values of all the suffixes from those of a subset.

	Assume $x[n - 1]$ is lexicographically the smallest character in $\Sigma$ and appears in $x[0, n)$ only once, we introduce the following notations and symbols for description clarity.

	{\em L-type/S-type/LMS character and suffix.} All the characters in $x$ are classified into three different types, namely L-type, S-type and LMS. In details, $x[i]$ is S-type if (1) $i = n - 1$ or (2) $x[i] < x[i + 1]$ or (3) $x[i] = x[i+1]$ and $x[i+1]$ is S-type; otherwise, $x[i]$ is L-type. In particular, if $x[i - 1]$ and $x[i]$ are respectively L-type and S-type, then $x[i]$ is LMS. Furthermore, ${\sf suf}(i)$ is L-type, S-type or LMS if $x[i]$ is L-type, S-type or LMS, respectively.

	{\em L-type/S-type/LMS substring.} Suppose $x[j]$ is LMS and no characters in $x[i + 1, j - 1]$ are LMS, then substring $x[i, j]$ is an L-type, S-type or LMS substring if $x[i]$ is L-type, S-type or LMS, respectively.

	{\em Suffix buckets.} Suffixes in $sa$ are naturally grouped into multiple buckets, where each bucket occupies a contiguous interval of $sa$ and contains all the suffixes with an identical heading character. Besides, a bucket can be further divided into two parts, where the left/right part is a subbucket containing only L-type/S-type suffixes. For short, we use ${\sf sa\_bkt}(c)$ to denote the bucket containing all the suffixes with a heading character $c$, and ${\sf sa\_bkt_L}(c)/{\sf sa\_bkt_S}(c)$ to denote the left/right subbucket in ${\sf sa\_bkt}(c)$.

	{\em LCP buckets.} Suppose ${\sf sa\_bkt}(c)$ is a suffix bucket ranging from $sa[i]$ to $sa[j]$, then $lcp[i, j]$ is the corresponding LCP bucket, denoted by ${\sf lcp\_bkt}(c)$, and it can be further divided into two subbuckets ${\sf lcp\_bkt_L}(c)$ and ${\sf lcp\_bkt_S}(c)$ according to ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$.

	{\em Character repetition count.} Suppose $x[i] = x[i + 1] = \dots = x[j]$ and $x[j] \ne x[j + 1]$, then $j - i + 1$ is the number of successive repetitions for $x[i]$. Let ${\sf rep}(i)$ denote the value.

	{\em Inverse suffix array.} The inverse suffix array $isa$ satisfies that $isa[sa[i]] = i$ for $i \in [0, n)$.

	{\em Reduced string.} The reduced string $x_1$ records the names of LMS suffixes, where each name represents the relative order of the corresponding suffix among all the LMS ones.

	{\em Position array.} The position array $pa$ records the starting positions of LMS suffixes.

	{\em Type array.} The type array $t$ records the types of characters in $x$. Let $t[i] = 0$ or $1$ if $x[i]$ is L-type or S-type, respectively.

	In addition, we denote the suffix and LCP arrays of LMS suffixes by $sa_{\sf LMS}$ and $lcp_{\sf LMS}$.

@@

	\begin{Corollary} \label{corollary:2}
	Given that $sa$ is correct, then $lcp$ is correct with a high probability if and only if these conditions are satisfied for all $c \in \Sigma$:

		(1) $lcp_{\sf LMS}$ is correct.

		(2) The fingerprints of $lcp_{\sf L1}(c)$ and $lcp_{\sf L2}(c)$ are equal.

		(3) The fingerprints of $lcp_{\sf S2}(c)$ and $lcp_{\sf S2}(c)$ are equal.

>>>>

While the study for efficient construction of suffix and LCP arrays is evolving, the programs implementing the proposed algorithms are commonly provided ``as is", with the purpose only for the performance evaluation experiemnts of the articles where they are reported. That is, these programms give no guarantee that they have correctly implemented the proposed algorithms. The programs for recently proprosed algorithms are becoming much more complicated than before, casuing more difficulties for program verifying and debugging\footnote{In our studies before, we have exprienced problems caused by bugs of the exisitng programs.}. As a common practice, a suffix or LCP array checker is also provided for verifying the correctness of a constructed array. For example, such a checker is provided in the software packages eSAIS \cite{xxx} (DC3, pScan and etc? more and better) for constructing suffix and/or LCP arrays. In addition to help avoid implementation bugs, a checker is also demanded for an array constructed by a probabilistic algorithm \cite{Bille2013}. In this case, the array is correct with a probability and hence must be verified by a checker to ensure its correctness.

@@

\SetKwProg{Fn}{Function}{}{}

@@

	%\SetAlgoNoLine
	\Fn{{\sf CheckByFP}($x$, $sa$, $lcp$, $n$)}{

@@

			\If{$ST_1.{\sf notEmpty}()$ {\sf and} $ST_1.{\sf top}().1st = i$}{~\label{alg:1:h}
				$e := ST_1.{\sf top}()$, $ST_1.{\sf pop}()$, $e.3rd := fp$, $ST_1'.{\sf push}(e)$
			}
			\Else{
				\Return false \hspace{5cm} // condition (1) is violated
			}~\label{alg:1:i}
			\While{$ST_2.{\sf notEmpty}()$ {\sf and} $ST_2.{\sf top}().1st = i$}{

@@

		\Return true
	}
\end{algorithm*}

@@

As will be seen from our experimental study in Section~\ref{sec:experiment}, Algorithm~\ref{alg:1} is quite space consuming, its peak disk use is 40 bytes per input character. In this section, we describe an alterative based on the induced-sorting principle, where the algorithm designed by this method only takes half space on real-world datasets in comparison with Algorithm~\ref{alg:1}.

\subsection{Preliminaries} \label{sec:method2:preliminaries}

Before our presentation, we first introduce some notations for description convenience.

{\em L-type/S-type/LMS character and suffix.} All the characters in $x$ are classified into three types, namely L-, S- and S*-type. Detailedly, $x[i]$ is L-type if (1) $i = n - 1$ or (2) $x[i] > x[i + 1]$ or (3) $x[i] = x[i + 1]$ and $x[i + 1]$ is L-type; otherwise, $x[i]$ is S-type. Further, if $x[i]$ and $x[i + 1]$ are S- and L-type respectively, then $x[i]$ is also an S*-type character. Moreover, the type of a suffix is the same as that of its heading character.

{\em Buckets.} Suppose $sa$ is correct, then suffixes in $sa$ are naturally partitioned into multiple buckets and those with an identical heading character are grouped into one bucket occupying a contiguous interval. Further, a bucket can be divided into two parts, where the left and right part contain L- and S-type suffixes, respectively. For short, we use ${\sf sa\_bkt}(c)$ to denote the bucket storing suffixes starting with $c$ and ${\sf sa\_bkt_L}(c)/{\sf sa\_bkt_S}(c)$ to denote its left/right sub-bucket. Accordingly, $lcp$ can be decomposed into multiple buckets as well, where ${\sf lcp\_bkt}(c)/{\sf lcp\_bkt_L}(c)/{\sf lcp\_bkt_S}(c)$ store the LCP-values of suffixes in ${\sf sa\_bkt}(c)/{\sf sa\_bkt_L}(c)/{\sf sa\_bkt_S}(c)$ and their left lexical neighbors.

{\em $sa^*$ and $lcp^*$.} Suppose the total number of S*-type suffixes in $x$ is $n_1$, then $sa^*[0, n_1)$ and $lcp^*[0, n_1)$ indicate their lexical order and the LCP-values, where $x[sa^*[i]]$ is the heading character of the $(i + 1)$-th smallest S*-type suffix.

\subsection{Idea} \label{sec:method2:idea}

The induced sorting~(IS) principle has been successfully used to design algorithms for building suffix and LCP arrays in both internal and external memory. These algorithms mainly consist of a reduction phase for computing $sa^*$ and $lcp^*$ followed by an induction phase for inducing $sa$ and $lcp$ from $sa^*$ and $lcp^*$. In other words, suppose $sa^*$ and $lcp^*$ are already known, we can directly compute the suffix and LCP arrays by calling the inducing process of an existing IS-based construction algorithm. This enlightens us to use the following conditions for verification:

\begin{Lemma} \label{lemma:2}
Both $sa[0, n)$ and $lcp[0, n)$ are correct if and only if the conditions below are satisfied:

\begin{enumerate}[(1)]
	\item
	$sa^*$ and $lcp^*$ are both correct.
	\item
	$sa = sa'$ and $lcp = lcp'$, where $sa'$ and $lcp'$ are induced from $sa^*$ and $lcp^*$ by calling the inducing procedure of an existing IS-based construction algorithm.
\end{enumerate}
\end{Lemma}

Similar to Section~\ref{sec:method1:idea}, we draw the conclusion in Corollary~\ref{corollary:2} by using the fingerprinting techinque to check condition (1) in Lemma~\ref{lemma:2}.

\begin{Corollary} \label{corollary:2}
Both $sa[0, n)$ and $lcp[0, n)$ are correct with a high probability given the following conditions, for $i \in [0, n)$, $j, k \in [1, n_1)$ and $j \ne k$:

\begin{enumerate}[(1)]
	\item 
	$sa^*[j] \ne sa^*[k]$. 
	\item
	${\sf fp}(sa^*[j], sa^*[j] + lcp^*[j] - 1) = {\sf fp}(sa^*[k - 1], sa^*[k - 1] + lcp^*[k] - 1)$.
	\item
	$x[sa^*[j] + lcp^*[j]] > x[sa^*[k - 1] + lcp^*[k]]$.
	\item 
	$sa[i] = sa'[i]$ and $lcp[i] = lcp'[i]$, where $sa'$ and $lcp'$ are induced from $sa^*$ and $lcp^*$ by calling the inducing procedure of an existing IS-based construction algorithm.
\end{enumerate}

\subsection{Algorithm} \label{sec:method2:algorithm}

The algorithm for checking Corollary~\ref{corollary:2} 	
The following procedure can b
For checking the conditions in \ref{corollary:2}, we

	\caption{The Algorithm for checking $sa^*$ and $lcp^*$.}

	\KwIn{$x$, $sa$, $lcp$, $n$}

	$ST_1$ := $[(sa[i], i) | i \in [0, n)]$ \label{alg:2:a}

	sort tuples in $ST_1$ by 1st component \label{alg:2:b}

	$j := 0$

	\For{$i \in (n, 0]$}{

		$e := ST_1.{\sf top}()$, $ST_1.{\sf pop}()$

		\If{$x[i]$ is S*-type}{

			$e.1st := j$, $j := j + 1$, $ST_2.{\sf push}(e)$
		}
	}

	sort tuples in $ST_2$ by 2nd component

	$i := 0, j := 0, lcp_{min} := max\_val$ \hspace{1cm} // use $lcp_{min}$ to trace the LCP-value of two successive S*-type suffixes

	\While{$ST_2.{\sf NotEmpty()}$}{

		$e := ST_2.{\sf top}()$, $ST_2.{\sf pop}()$

		\While {true} {

			$i:= i + 1$, $lcp_{min} := {\sf min}(lcp_{min}, lcp[i])$

			\If {$e.2nd = i$} {

				$sa^*[j] := e.1st$, $lcp^*[j] := lcp_{min}$, $j := j + 1$

				\textbf{break}
			}
		}

		$lcp_{min} := max\_val$
	}

	CheckByFP\_Modified($x, sa^*, lcp^*, n_1$)

@@
